Spatial Reasoning and Data Displays

### 2015-03-31 Paper/InfoVis2015/TestingVisualAptitude.Rnw
labels fixed

%% Uncomment below to include a teaser figure.
   \teaser{
      \centering
      \begin{minipage}[c]{0.4\textwidth}
         \subfloat[][``Which of these plots is the most different from the others?"]{\label{fig:lineup} \includegraphics[height=\textwidth]{lineup.png}}
        %
        \end{minipage}
        \begin{minipage}[c]{.4\textwidth}
         \subfloat[][Biplots]{\label{fig:biplots}
          \includegraphics[height=.8\textwidth]{Figure/fig-biplots-pca5-1}\hfill\newline
          \includegraphics[height=.8\textwidth]{Figure/fig-biplots-pca5-2}
         }
          \end{minipage}
    \caption{On the left a sample lineup of boxplots is shown. Participants are asked to identify the most different plot.  %Participants are instructed to choose the plot which appears most different from the others. 
    Here, panel 4 is the target plot, because the two groups have a large difference in medians.\\
    The plots on the right show biplots of principal components 2-5 with observations. The lineup task appears to be most similar to the figure classification task, based on the plot of PC2 vs. PC3.}
  }

@@

  qplot(u[,1], u[,2], data=u, colour=I("grey50"), shape=I(1), size=I(3)) + theme_bw() + 
    theme(aspect.ratio=1) + coord_equal() + 
  geom_segment(aes(x=0, y=0, xend=v[,1], yend=v[,2], colour=segColor), data=v, arrow=arrow(length = unit(0.3, "cm"))) + 
    xlab(sprintf("PC%d",choices[1])) + ylab(sprintf("PC%d",choices[2])) +
    geom_text(aes(x=v[,1], y=v[,2], label=label, colour=segColor, 
                  hjust=0.15*(0.5-sign(v[,1])), vjust=0.5-sign(v[,2])), data=v) +
    scale_colour_identity() + theme(aspect.ratio=1)
}
@

@@

Between cognitive tasks, participants were also asked to complete three blocks of 20 lineups each, assembled from previous studies~\cite{hofmann2012graphical,majumder2013validation}. Participants were given 5 minutes to complete each block of 20 lineups. Figure~\ref{fig:lineup} shows a sample lineup of box plots.

@@

<<biplot-pca4, fig.cap="", fig.width=4.5, fig.height=4.5, warning=F, message=F, out.width='.45\\linewidth', comment="|", fig.cap="Biplots of principal components 1-4 with observations (same as figure 1(b)). Principal component analysis was performed on the four cognitive tests used to understand the association between the cognitive skills required for these tests and the skills required for the lineup protocol.  \\label{fig:biplots4}", echo=F, fig.env="figure">>=
#par(mar=c(4,4,.1,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
dimnames(pca2$rotation)[[1]] <- c("card rotation", "figure classification", "paper folding", "visual search")

ggbiplot(pca2, choices=1:2) + xlim(c(-2.5,2.5)) +  ylim(c(-2.5,2.5))
ggbiplot(pca2, choices=3:4) + xlim(c(-3,3)) + ylim(c(-2.5,2.5))
@
Figure~\ref{fig:biplots4} shows that the first PC does not differentiate between any of the tasks; it might be best understood as a general aptitude factor. All of the remaining principal components distinguish between the cognitive tasks; we can utilize this separation by incorporating the lineup task and examining changes in the principal components.

@@

% <<biplots-pca5, fig.cap="", fig.width=4.5, fig.height=4.5, warning=F, message=F, out.width='.45\\linewidth', comment="|", fig.cap="Biplots of principal components 2-5 with observations. The lineup task appears to be most similar to the figure classification task, based on the plot of PC2 vs. PC3.  \\label{fig:biplots5}", echo=F, fig.env="figure">>=
% par(mar=c(4,4,.1,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
% biplot(pca, choices=2:3, pc.biplot=T, cex=c(.5, 1), xlim=c(-2.5,2.5), ylim=c(-2.5,2.5))
% biplot(pca, choices=4:5, pc.biplot=T, cex=c(.5, 1), xlim=c(-2,2), ylim=c(-2,2))
% @

<<biplots-pca5, fig.cap="", fig.width=4.5, fig.height=4.5, warning=F, message=F, out.width='.45\\linewidth', comment="|", fig.cap="Biplots of principal components 2-5 with observations. The lineup task appears to be most similar to the figure classification task, based on the plot of PC2 vs. PC3.  \\label{fig:biplots5}", echo=F, fig.env="figure">>=

dimnames(pca$rotation)[[1]] <- c("lineup", "card rotation", "figure classification", "folding", "visual search")
ggbiplot(pca, choices=2:3, segColor=c("#CC3333", rep("black", 4))) + xlim(c(-2.6,2.6)) +  ylim(c(-2.6,2.6))
ggbiplot(pca, choices=4:5, segColor=c("#CC3333", rep("black", 4)))
@
Biplots \cite{biplots:gabriel} of the remaining components are provided in figure~\ref{fig:biplots5}. Figure classification is strongly related to lineups (PC2, PC3). Performance on the visual search task  is also related to lineup performance (PC3). These two components highlight the shared demands of the lineup task and the figure classification task: participants must establish categories from provided stimuli and then classify the stimuli accordingly.

>>>>

%% Uncomment below to include a teaser figure.
   \teaser{
      \centering
      \hfill
      \begin{minipage}[c]{0.4\textwidth}
          \includegraphics[height=\textwidth]{lineup.png}
          (a) ``Which of these plots is the most different from the others?"
        %
        \end{minipage}
      %  \hfill
        \begin{minipage}[c]{.4\textwidth}
        \centering
          \includegraphics[height=.5\textwidth]{Figure/fig-biplots-pca5-1}\hfill\newline
          \includegraphics[height=.5\textwidth]{Figure/fig-biplots-pca5-2}\hfill\newline

        (b) Biplots of visuo-spatial tests and lineup task.
\end{minipage}
\hfill
    \caption{\label{fig:lineup}On the left a sample lineup of boxplots is shown. Participants are asked to identify the most different plot.  %Participants are instructed to choose the plot which appears most different from the others. 
    Panel 4 is the target plot, because the two groups have a large difference in medians.\\
    The biplots on the right show principal components 2-5 with observations from the user study. The lineup task appears to be most similar to the figure classification task.}
  }

@@

  qplot(u[,1], u[,2], data=u, colour=I("grey50"), shape=I(1), size=I(3)) + theme_bw() + 
    theme(aspect.ratio=1) + coord_equal() + 
  geom_segment(aes(x=0, y=0, xend=v[,1], yend=v[,2], colour=segColor), data=v, arrow=arrow(length = unit(0.3, "cm"))) + 
    xlab(sprintf("PC%d",choices[1])) + ylab(sprintf("PC%d",choices[2])) +
    geom_text(aes(x=v[,1], y=v[,2], label=label, colour=segColor, 
                  hjust=(0.5-sign(v[,1])/2), vjust=(0.5-sign(v[,2])/2) ), data=v) +
    scale_colour_identity() + theme(aspect.ratio=1)
}
@

@@

Between cognitive tasks, participants were also asked to complete three blocks of 20 lineups each, assembled from previous studies~\cite{hofmann2012graphical,majumder2013validation}. Participants were given 5 minutes to complete each block of 20 lineups. Figure~\ref{fig:lineup}(a) shows a sample lineup of box plots.

@@

<<biplot-pca4, fig.cap="", fig.width=5.5, fig.height=5.5, warning=F, message=F, out.width='.45\\linewidth', comment="|", fig.cap="Biplots of principal components 1-4 with observations (same as figure 1(b)). Principal component analysis was performed on the four cognitive tests used to understand the association between the cognitive skills required for these tests and the skills required for the lineup protocol.  \\label{fig:biplots4}", echo=F, fig.env="figure">>=
#par(mar=c(4,4,.1,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
dimnames(pca2$rotation)[[1]] <- c("card\nrotation", "figure\nclassification", "paper\nfolding", "visual\nsearch")

ggbiplot(pca2, choices=1:2) + xlim(c(-2.5,4.1)) +  ylim(c(-2.5,2.5))
ggbiplot(pca2, choices=3:4) + xlim(c(-3,3.5)) + ylim(c(-3,3))
@
Figure~\ref{fig:biplots4} shows that the first PC does not differentiate between any of the tasks; it might be best understood as a general aptitude factor. All of the remaining principal components distinguish between the cognitive tasks; we can utilize this separation by incorporating the lineup task and examining changes in the principal components.

@@

<<biplots-pca5, fig.cap="", fig.width=5.5, fig.height=5.5, warning=F, message=F, out.width='.45\\linewidth', comment="|", fig.cap="Biplots of principal components 2-5 with observations. The lineup task appears to be most similar to the figure classification task, based on the plot of PC2 vs. PC3. (These are the same biplots as shown in figure \\ref{fig:lineup}b)  \\label{fig:biplots5}", echo=F, fig.env="figure">>=

dimnames(pca$rotation)[[1]] <- c("lineup", "card\nrotation", "figure\nclassification", "paper\nfolding", "visual\nsearch")
ggbiplot(pca, choices=2:3, segColor=c("#CC3333", rep("black", 4))) + xlim(c(-2.4,2.8)) +  ylim(c(-2.6,2.6))
ggbiplot(pca, choices=4:5, segColor=c("#CC3333", rep("black", 4))) + xlim(c(-2.8,2.8)) + ylim(c(-2.8,2.8))
@
Biplots \cite{biplots:gabriel} of the remaining components are provided in figure~\ref{fig:biplots5}. Figure classification is strongly related to lineups (PC2, PC3). Performance on the visual search task  is also related to lineup performance (PC3). These two components highlight the shared demands of the lineup task and the figure classification task: participants must establish categories from provided stimuli and then classify the stimuli accordingly.

### 2015-03-31 Paper/InfoVis2015/TestingVisualAptitude.tex
labels fixed

%% Uncomment below to include a teaser figure.
   \teaser{
      \centering
      \begin{minipage}[c]{0.4\textwidth}
         \subfloat[][``Which of these plots is the most different from the others?"]{\label{fig:lineup} \includegraphics[height=\textwidth]{lineup.png}}
        %
        \end{minipage}
        \begin{minipage}[c]{.1\textwidth}
         \subfloat[][Biplots]{\label{fig:biplots}
          \includegraphics[height=.8\textwidth]{Figure/fig-biplots-pca5-1}\hfill\newline
          \includegraphics[height=.8\textwidth]{Figure/fig-biplots-pca5-2}
         }
          \end{minipage}
    \caption{On the left a sample lineup of boxplots is shown. Participants are asked to identify the most different plot.  %Participants are instructed to choose the plot which appears most different from the others. 
    Here, panel 4 is the target plot, because the two groups have a large difference in medians.\\
    The plots on the right show biplots of principal components 2-5 with observations. The lineup task appears to be most similar to the figure classification task, based on the plot of PC2 vs. PC3.}
  }

@@

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Mon Mar 30 19:13:11 2015
\begin{table}[htb]
\centering
\caption{Importance of principal components in an analysis of four tests of spatial ability: figure classification, paper folding, card rotation, and visual search.\label{tab:PCAvariance4}} 
{\footnotesize
\begin{tabular}{rrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 \\ 
  \hline
Standard deviation & 1.61 & 0.81 & 0.73 & 0.49 \\ 
  Proportion of Variance & 0.64 & 0.16 & 0.13 & 0.06 \\ 
  Cumulative Proportion & 0.64 & 0.81 & 0.94 & 1.00 \\ 
   \hline
\end{tabular}
}
\end{table}

@@

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Mon Mar 30 19:13:11 2015
\begin{table}[htb]
\centering
\caption{Rotation matrix for principal component analysis of the four cognitive tests (visual search, paper folding, card rotation, figure classification).\label{tab:PCArotation4}} 
\begin{tabular}{rrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 \\ 
  \hline
card.rot & 0.55 & -0.19 & -0.38 & 0.72 \\ 
  fig.class & 0.46 & 0.58 & 0.66 & 0.14 \\ 
  folding & 0.52 & 0.33 & -0.53 & -0.59 \\ 
  vis.search & 0.46 & -0.72 & 0.38 & -0.34 \\ 
   \hline
\end{tabular}
\end{table}

@@

\caption{Biplots of principal components 1-4 with observations. Principal component analysis was performed on the four cognitive tests used to understand the association between the cognitive skills required for these tests and the skills required for the lineup protocol.  \label{fig:biplots4}\label{fig:biplot-pca4}}
\end{figure}

@@

Next, we add the lineup results to our principal components analysis. The results of the principal component analysis of the four cognitive tests and lineup results are similar to the results of the cognitive tests alone; this suggests that lineups require at least some of the skills measured by the other cognitive tests. Table~\ref{tab:PCAvariance5} shows the importance of each principal component; from the distribution of the variance components, we see that the lineup test spans an additional dimension within the space of the four established tests. 
% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Mon Mar 30 19:13:11 2015
\begin{table}[ht]
\centering
\caption{Importance of principal components, analyzing all five tests.\label{tab:PCAvariance5}} 
{\footnotesize
\begin{tabular}{rrrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 \\ 
  \hline
Standard deviation & 1.73 & 0.84 & 0.75 & 0.70 & 0.48 \\ 
  Proportion of Variance & 0.60 & 0.14 & 0.11 & 0.10 & 0.05 \\ 
  Cumulative Proportion & 0.60 & 0.74 & 0.85 & 0.95 & 1.00 \\ 
   \hline
\end{tabular}
}
\end{table}

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Mon Mar 30 19:13:11 2015
\begin{table}[ht]
\centering
\caption{PCA Rotation matrix for all five tests. The first principal component is essentially an average of all five tests. 
             \label{tab:PCArotation5}} 
\begin{tabular}{rrrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 \\ 
  \hline
lineup & 0.42 & 0.49 & -0.46 & 0.60 & -0.10 \\ 
  card.rot & 0.50 & -0.30 & 0.28 & 0.23 & 0.73 \\ 
  fig.class & 0.43 & 0.45 & -0.15 & -0.75 & 0.18 \\ 
  folding & 0.47 & 0.07 & 0.68 & 0.04 & -0.56 \\ 
  vis.search & 0.41 & -0.69 & -0.48 & -0.15 & -0.33 \\ 
   \hline
\end{tabular}
\end{table}

@@

% <<biplots-pca5, fig.cap="", fig.width=4.5, fig.height=4.5, warning=F, message=F, out.width='.45\\linewidth', comment="|", fig.cap="Biplots of principal components 2-5 with observations. The lineup task appears to be most similar to the figure classification task, based on the plot of PC2 vs. PC3.  \\label{fig:biplots5}", echo=F, fig.env="figure">>=
% par(mar=c(4,4,.1,.1),cex.lab=.95,cex.axis=.9,mgp=c(2,.7,0),tcl=-.3)
% biplot(pca, choices=2:3, pc.biplot=T, cex=c(.5, 1), xlim=c(-2.5,2.5), ylim=c(-2.5,2.5))
% biplot(pca, choices=4:5, pc.biplot=T, cex=c(.5, 1), xlim=c(-2,2), ylim=c(-2,2))
% @

@@

\caption{Biplots of principal components 2-5 with observations. The lineup task appears to be most similar to the figure classification task, based on the plot of PC2 vs. PC3.  \label{fig:biplots5}\label{fig:biplots-pca5}}
\end{figure}

@@

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Mon Mar 30 19:13:12 2015
\begin{table}[ht]
\centering
\caption{Estimates and significances of a linear model of lineup scores.} 
\label{tab:m1}
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & 14.1192 & 1.9149 & 7.37 & 0.0000 \\ 
  PC1 & 1.7672 & 0.5230 & 3.38 & 0.0018 \\ 
  MATH & 2.1246 & 0.8732 & 2.43 & 0.0202 \\ 
   \hline
\end{tabular}
\end{table}

>>>>

%% Uncomment below to include a teaser figure.
   \teaser{
      \centering
      \hfill
      \begin{minipage}[c]{0.4\textwidth}
          \includegraphics[height=\textwidth]{lineup.png}
          (a) ``Which of these plots is the most different from the others?"
        %
        \end{minipage}
      %  \hfill
        \begin{minipage}[c]{.4\textwidth}
        \centering
          \includegraphics[height=.5\textwidth]{Figure/fig-biplots-pca5-1}\hfill\newline
          \includegraphics[height=.5\textwidth]{Figure/fig-biplots-pca5-2}\hfill\newline

        (b) Biplots of visuo-spatial tests and lineup task.
\end{minipage}
\hfill
    \caption{\label{fig:lineup}On the left a sample lineup of boxplots is shown. Participants are asked to identify the most different plot.  %Participants are instructed to choose the plot which appears most different from the others. 
    Panel 4 is the target plot, because the two groups have a large difference in medians.\\
    The biplots on the right show principal components 2-5 with observations from the user study. The lineup task appears to be most similar to the figure classification task.}
  }

@@

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Mon Mar 30 23:08:34 2015
\begin{table}[htb]
\centering
\caption{Importance of principal components in an analysis of four tests of spatial ability: figure classification, paper folding, card rotation, and visual search.\label{tab:PCAvariance4}} 
{\footnotesize
\begin{tabular}{rrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 \\ 
  \hline
Standard deviation & 1.61 & 0.81 & 0.73 & 0.49 \\ 
  Proportion of Variance & 0.64 & 0.16 & 0.13 & 0.06 \\ 
  Cumulative Proportion & 0.64 & 0.81 & 0.94 & 1.00 \\ 
   \hline
\end{tabular}
}
\end{table}

@@

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Mon Mar 30 23:08:34 2015
\begin{table}[htb]
\centering
\caption{Rotation matrix for principal component analysis of the four cognitive tests (visual search, paper folding, card rotation, figure classification).\label{tab:PCArotation4}} 
\begin{tabular}{rrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 \\ 
  \hline
card.rot & 0.55 & -0.19 & -0.38 & 0.72 \\ 
  fig.class & 0.46 & 0.58 & 0.66 & 0.14 \\ 
  folding & 0.52 & 0.33 & -0.53 & -0.59 \\ 
  vis.search & 0.46 & -0.72 & 0.38 & -0.34 \\ 
   \hline
\end{tabular}
\end{table}

@@

\caption{Biplots of principal components 1-4 with observations (same as figure 1(b)). Principal component analysis was performed on the four cognitive tests used to understand the association between the cognitive skills required for these tests and the skills required for the lineup protocol.  \label{fig:biplots4}\label{fig:biplot-pca4}}
\end{figure}

@@

Next, we add the lineup results to our principal components analysis. The results of the principal component analysis of the four cognitive tests and lineup results are similar to the results of the cognitive tests alone; this suggests that lineups require at least some of the skills measured by the other cognitive tests. Table~\ref{tab:PCAvariance5} shows the importance of each principal component; from the distribution of the variance components, we see that the lineup test spans an additional dimension within the space of the four established tests. 
% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Mon Mar 30 23:08:35 2015
\begin{table}[ht]
\centering
\caption{Importance of principal components, analyzing all five tests.\label{tab:PCAvariance5}} 
{\footnotesize
\begin{tabular}{rrrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 \\ 
  \hline
Standard deviation & 1.73 & 0.84 & 0.75 & 0.70 & 0.48 \\ 
  Proportion of Variance & 0.60 & 0.14 & 0.11 & 0.10 & 0.05 \\ 
  Cumulative Proportion & 0.60 & 0.74 & 0.85 & 0.95 & 1.00 \\ 
   \hline
\end{tabular}
}
\end{table}

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Mon Mar 30 23:08:35 2015
\begin{table}[ht]
\centering
\caption{PCA Rotation matrix for all five tests. The first principal component is essentially an average of all five tests. 
             \label{tab:PCArotation5}} 
\begin{tabular}{rrrrrr}
  \hline
 & PC1 & PC2 & PC3 & PC4 & PC5 \\ 
  \hline
lineup & 0.42 & 0.49 & -0.46 & 0.60 & -0.10 \\ 
  card.rot & 0.50 & -0.30 & 0.28 & 0.23 & 0.73 \\ 
  fig.class & 0.43 & 0.45 & -0.15 & -0.75 & 0.18 \\ 
  folding & 0.47 & 0.07 & 0.68 & 0.04 & -0.56 \\ 
  vis.search & 0.41 & -0.69 & -0.48 & -0.15 & -0.33 \\ 
   \hline
\end{tabular}
\end{table}

@@

@@

\caption{Biplots of principal components 2-5 with observations. The lineup task appears to be most similar to the figure classification task, based on the plot of PC2 vs. PC3. (These are the same biplots as shown in figure \ref{fig:lineup}b)  \label{fig:biplots5}\label{fig:biplots-pca5}}
\end{figure}

@@

% latex table generated in R 3.1.1 by xtable 1.7-3 package
% Mon Mar 30 23:08:36 2015
\begin{table}[ht]
\centering
\caption{Estimates and significances of a linear model of lineup scores.} 
\label{tab:m1}
\begin{tabular}{rrrrr}
  \hline
 & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
  \hline
(Intercept) & 14.1192 & 1.9149 & 7.37 & 0.0000 \\ 
  PC1 & 1.7672 & 0.5230 & 3.38 & 0.0018 \\ 
  MATH & 2.1246 & 0.8732 & 2.43 & 0.0202 \\ 
   \hline
\end{tabular}
\end{table}
